---
title: "GPP Summary"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```


```{r makeDFs, include=FALSE}
library(raster)
library(ggplot2)
library(dplyr)
library(palr)
library(rasterVis)

meanAnnualGPPfile = "D:/work/proj/npp/20171107/mean_annual/gpp_1080x2160_mean_annual_g.tif"
distFile = "D:/work/proj/npp/20171107/mean_annual/land_water_dist_geo.tif"
meanAnnualGPPfileFresh = "D:/work/proj/npp/20171107/freshwater/StreamProductivity.csv"
width = 5


# read in freshwater
ggppFresh = read.csv(meanAnnualGPPfileFresh)

# read in the file and get the number of rows
ggpp = raster(meanAnnualGPPfile)
distR = raster(distFile)
nRow = nrow(ggpp)

# bind the rows into latitude bands
fromRow = seq(1, nRow, 6*width)
toRow = seq(6*width, nRow, 6*width)
length(fromRow) == length(toRow)
lats = seq(90, -90, -180/nRow)
fromLat = lats[fromRow]
toLat = c(fromLat[2:length(fromLat)], -90)
label = paste0(fromLat,' : ',toLat)

hemi = midLat = vector()
for(i in 1:length(toLat)){
  mid = ((fromLat[i]-toLat[i])/2)+toLat[i]
  midLat = c(midLat,mid)
  hem = 'North'
  if(mid < 0){
    hem = 'South'
  }
  hemi = c(hemi, hem)
}



fromLatAbs = abs(fromLat)
toLatAbs = abs(toLat)
midLatAbs = abs(midLat)
labelAbs = paste0(fromLatAbs,' : ',toLatAbs)

latInfo = data.frame(fromRow, toRow, fromLat, toLat, midLat, label,
                     fromLatAbs, toLatAbs, midLatAbs, labelAbs, hemi)



# fresh water df
dfSalt = dfFresh = data.frame(mean=NA, stdev=NA, median=NA, q1=NA, q3=NA, 
                              fromLat=latInfo$fromLat, toLat=latInfo$toLat, midLat=latInfo$midLat, label=latInfo$label,
                              fromLatAbs=latInfo$fromLatAbs, toLatAbs=latInfo$toLatAbs, midLatAbs=latInfo$midLatAbs, labelAbs=latInfo$labelAbs,
                              hemi = latInfo$hemi)
dfFresh$water = 'Fresh'
dfSalt$water = 'Salt'


for(i in 1:nrow(dfFresh)){
  theseRows = filter(ggppFresh, Lat_Deg <= dfFresh$fromLat[i] & Lat_Deg > dfFresh$toLat[i])
  mean_ = mean(theseRows$Annual_GPP, na.rm=T)
  sd_ = sd(theseRows$Annual_GPP, na.rm=T)
  medq = summary(theseRows$Annual_GPP, na.rm=T)
  if(sum(!is.finite(c(mean_, sd_))) < 2){
    dfFresh$mean[i] = mean_
    dfFresh$stdev[i] = sd_
    dfFresh$median[i] = medq[3]
    dfFresh$q1[i] = medq[2]
    dfFresh$q3[i] = medq[5]
  }
}


# bring in the freshwater
# test = mutate(ggppFresh, label=cut(Lat_Deg, seq(90, -90, -width), rev(paste0(fromLat,' - ',toLat)))) %>%
#   group_by(label) %>% 
#   summarise(mean = mean(Annual_GPP, na.rm=T), stdev=sd(Annual_GPP, na.rm=T)) %>%
#   mutate(water = "Fresh")


# get salt water data
for(i in 1:nrow(latInfo)){
  print(paste0(latInfo$fromLat[i],'-',latInfo$toLat[i]))
  ggppBin = ggpp[latInfo$fromRow[i]:latInfo$toRow[i],]
  medq = summary(ggppBin, na.rm=T)
  dfSalt$mean[i] = mean(ggppBin, na.rm=T)
  dfSalt$stdev[i] = sd(ggppBin, na.rm=T)
  dfSalt$median[i] = medq[3]
  dfSalt$q1[i] = medq[2]
  dfSalt$q3[i] = medq[5]
}

# plot gpp as function of latitude
# dfSalt = na.omit(dfSalt)
# latP = ggplot(dfSalt, aes(x=midLat, y=mean))+
#   geom_ribbon(aes(ymin = mean - stdev, ymax = mean + stdev), fill = "grey70") +
#   geom_line()+
#   ylab("Carbon (g/m^2/year)")+
#   xlab("Latitude (degrees)")+
#   theme_bw()
# 
# latP



# combine fresh and salt
combined = rbind(dfFresh, dfSalt)

combined = mutate(combined, class=paste(hemi, water))

# set label as factor
combined$label = as.factor(combined$label)
combined$label = factor(combined$label, levels = rev(latInfo$label))

combined$labelAbs = as.factor(combined$labelAbs)
combined$labelAbs = factor(combined$label, levels = rev(latInfo$labelAbs))

combined$class = as.factor(combined$class)
combined$hemi = as.factor(combined$hemi)
combined$water = as.factor(combined$water)



# plot gpp as function of latitude LINES MEAN
# latP = ggplot(combined, aes(x=midLat, y=mean, color=water))+
#   #geom_ribbon(aes(ymin = mean - stdev, ymax = mean + stdev, fill=water), alpha=0.1) +
#   geom_point()+
#   geom_line(size=2)+
#   ylab("Carbon (g/m^2/year)")+
#   xlab("Latitude (degrees)")+
#   theme_bw()
# 
# latP

dodge = position_dodge(3)
```

Gross found that migration to freshwater is more frequent in tropical latitudes, where freshwater productivity is greater than that of oceans. He found three important patterns in his data.

1. Freshwater productivity is higher than ocean productivity at low latitudes.
2. Ocean productivity is higher than freshwater at high latitudes.
3. Ocean productivity is generally higher at high latitudes and low at equatorial latitudes.  The opposite is true for freshwater.

We took a look at updated data to see if these patterns held. We found that they do not. Notably, freshwater productivity is always lower than ocean productivity. (Figures 1, 2). One exception occurs at 50 degrees latitude, but that is driven by a single point that is much, much larger than any other observation, and we may want to remove it from the analysis.

##### Fig 1. Mean & Stdev GPP by latitude w/ hemispheres on x-axis
```{r fig1, fig.height = 4, out.width = '100%', echo=FALSE}
latP = ggplot(combined, aes(x=midLat, y=mean, color=water))+
  geom_errorbar(aes(ymin = mean - stdev, ymax = mean + stdev, color=water), width=0.1, position = dodge, alpha=0.8) +
  geom_point(size=2, position = dodge, alpha=0.8)+
  ylab("Carbon (g/m^2/year)")+
  xlab("Latitude (degrees)")+
  guides(color=guide_legend(title="Water Type"))+
  theme_bw()
latP
```

##### Fig 2. Median & interquartile GPP by latitude w/ hemispheres on x-axis
```{r fig2, fig.height = 4, out.width = '100%', echo=FALSE}
latP = ggplot(combined, aes(x=midLat, y=median, color=water))+
  geom_errorbar(aes(ymin = q1, ymax = q3, color=water), width=0.1, position = dodge, alpha=0.8) +
  geom_point(size=2, position = dodge, alpha=0.8)+
  ylab("Carbon (g/m^2/year)")+
  xlab("Latitude (degrees)")+
  guides(color=guide_legend(title="Water Type"))+
  theme_bw()
latP
```

The patterns that ocean productivity increases with latitude, while freshwater productivity decreases with latitude are more complicated.

Freshwater productivity does not show any meaningful trend. The sample size for freshwater is very small (140 data points), so there is no way to know if the lack of trend is because of data or because there is no trend. There is data to show that freshwater productivity is heavily influenced by pollution (Finlay 2011), stream size (McTammany et al. , Finlay 2011), season (Minshall et al. 1992), and watershed area (Finlay 2011). It is likely that even if there is a decreasing trend with latitude, it would be hard to detect without accounting for these other important factors (which we can't do very well because our data size is so small).

Ocean productivity does not increase with latitude in a linear (or simple) way. Starting at the equator, ocean productivity generally decreases as latitude increases, but then starts to increase at 25 degrees latitude. Both northern and southern hemispheres follow the same pattern until 25 degrees, but then their patterns diverge significantly (Figures 3, 4). At northern latitudes, productivity continues to increase to 55 degrees latitude, then declines. This pattern is shown in Gross' Figure 1B. The southern hemisphere productivity declines sharply after -30 degrees latitude, but then shows a slight increase around -77 degrees. This is really important because Gross 'folds' his data by latitude, but the productivity patterns are very different. 


<!--
## GPP by Latitude w/ hemispheres included in Group factor

#### Mean & Stdev GPP
```{r fig3, fig.height = 4, out.width = '100%', echo=FALSE}
#plot gpp as function of latitude POINTS MEAN
latP = ggplot(combined, aes(x=midLatAbs, y=mean, color=class))+
  geom_errorbar(aes(ymin = mean - stdev, ymax = mean + stdev, color=class), width=0.1, position = dodge, alpha=0.8) +
  geom_point(size=2, position = dodge, alpha=0.8)+
  ylab("Carbon (g/m^2/year)")+
  xlab("Latitude (degrees)")+
  guides(color=guide_legend(title="Group"))+
  theme_bw()
latP
```

#### Median & interquartile GPP 
```{r fig4, fig.height = 4, out.width = '100%', echo=FALSE}
latP = ggplot(combined, aes(x=midLatAbs, y=median, color=class))+
  geom_errorbar(aes(ymin = q1, ymax = q3, color=class), width=0.1, position = dodge, alpha=0.8) +
  geom_point(size=2, position = dodge, alpha=0.8)+
  ylab("Carbon (g/m^2/year)")+
  xlab("Latitude (degrees)")+
  guides(color=guide_legend(title="Group"))+
  theme_bw()
latP
```

## GPP by Latitude w/ hemispheres included in Group factor w/ regression lines

#### Mean & Stdev GPP
```{r fig5, fig.height = 4, out.width = '100%', echo=FALSE}
latP = ggplot(combined, aes(x=midLatAbs, y=mean, color=class))+
  geom_errorbar(aes(ymin = mean - stdev, ymax = mean + stdev, color=class), width=0.1, position = dodge, alpha=0.8) +
  geom_point(size=2, position = dodge, alpha=0.8)+
  geom_smooth(method='lm',formula=y~x, se=FALSE)+
  ylab("Carbon (g/m^2/year)")+
  xlab("Latitude (degrees)")+
  guides(color=guide_legend(title="Group"))+
  theme_bw()
latP
```

#### Median & interquartile GPP 
```{r fig6, fig.height = 4, out.width = '100%', echo=FALSE}
latP = ggplot(combined, aes(x=midLatAbs, y=median, color=class))+
  geom_errorbar(aes(ymin = q1, ymax = q3, color=class), width=0.1, position = dodge, alpha=0.8) +
  geom_point(size=2, position = dodge, alpha=0.8)+
  geom_smooth(method='lm',formula=y~x, se=FALSE)+
  ylab("Carbon (g/m^2/year)")+
  xlab("Latitude (degrees)")+
  guides(color=guide_legend(title="Group"))+
  theme_bw()
latP
```
-->



##### Fig 3. Mean & Stdev GPP by latitude w/ hemispheres as factor w/ facet on water type
```{r fig7, fig.height = 4, out.width = '100%', echo=FALSE}
latP = ggplot(combined, aes(x=midLatAbs, y=mean, color=hemi))+
  geom_errorbar(aes(ymin = mean - stdev, ymax = mean + stdev, color=hemi), width=0.1, position = dodge, alpha=0.8) +
  geom_point(size=2, position = dodge, alpha=0.8)+
  facet_grid(~water)+
  ylab("Carbon (g/m^2/year)")+
  xlab("Latitude (degrees)")+
  guides(color=guide_legend(title="Hemisphere"))+
  theme_bw()
latP
```

##### Fig. 4 Median & interquartile GPP by latitude w/ hemispheres as factor w/ facet on water type
```{r fig8, fig.height = 4, out.width = '100%', echo=FALSE}
latP = ggplot(combined, aes(x=midLatAbs, y=median, color=hemi))+
  geom_errorbar(aes(ymin = q1, ymax = q3, color=hemi), width=0.1, position = dodge, alpha=0.8) +
  geom_point(size=2, position = dodge, alpha=0.8)+
  facet_grid(~water)+
  ylab("Carbon (g/m^2/year)")+
  xlab("Latitude (degrees)")+
  guides(color=guide_legend(title="Hemisphere"))+
  theme_bw()
latP
```


<!--
## GPP by Latitude w/ hemispheres as factor w/ facet on water type w/ regression lines

#### Mean & Stdev GPP
```{r fig9, fig.height = 4, out.width = '100%', echo=FALSE}
latP = ggplot(combined, aes(x=midLatAbs, y=mean, color=hemi))+
  geom_errorbar(aes(ymin = mean - stdev, ymax = mean + stdev, color=hemi), width=0.1, position = dodge, alpha=0.8) +
  geom_point(size=2, position = dodge, alpha=0.8)+
  geom_smooth(method='lm',formula=y~x, se=FALSE)+
  facet_grid(~water)+
  ylab("Carbon (g/m^2/year)")+
  xlab("Latitude (degrees)")+
  guides(color=guide_legend(title="Hemisphere"))+
  theme_bw()
latP
```

#### Median & interquartile GPP
```{r fig10, fig.height = 4, out.width = '100%', echo=FALSE}
latP = ggplot(combined, aes(x=midLatAbs, y=median, color=hemi))+
  geom_errorbar(aes(ymin = q1, ymax = q3, color=hemi), width=0.1, position = dodge, alpha=0.8) +
  geom_point(size=2, position = dodge, alpha=0.8)+
  geom_smooth(method='lm',formula=y~x, se=FALSE)+
  facet_grid(~water)+
  ylab("Carbon (g/m^2/year)")+
  xlab("Latitude (degrees)")+
  guides(color=guide_legend(title="Hemisphere"))+
  theme_bw()
latP
```
-->






```{r makeHists, include=FALSE}
df = data.frame(values=NA, label=NA, labelAbs=NA, hemi=NA, midLatAbs=NA)

for(i in 1:nrow(latInfo)){
  print(latInfo$label[i])
  ggppBin = ggpp[latInfo$fromRow[i]:latInfo$toRow[i],]
  addDF = data.frame(values=ggppBin, label=latInfo$label[i], labelAbs=latInfo$labelAbs[i], hemi=latInfo$hemi[i], midLatAbs=latInfo$midLatAbs[i])
  df = rbind(df, addDF)
}

dfClean = na.omit(df)
nrow(dfClean)
dfClean$label = as.factor(dfClean$label)
dfClean$label = factor(dfClean$label, levels = rev(latInfo$label))

dfClean$labelAbs = as.factor(dfClean$labelAbs)
dfClean$labelAbs = factor(dfClean$labelAbs, levels = rev(latInfo$labelAbs))
```




Another really important point is that latitude is a very poor predictor of ocean productivity. Both the boxplots and map show this well (Figures 5, 6). In the box plots you can see that the data spread by latitude is enormous and the mean doesn't really capture what's going on with productivity in a meaningful way. 




##### Fig 5. Boxplots of ocean GPP w/ hemispheres included in x-axis
```{r fig11, fig.height = 8, out.width = '100%', echo=FALSE}
plt = ggplot(dfClean, aes(label, values))+
  geom_boxplot(outlier.alpha = 0.05)+
  ylab("Carbon (g/m^2/year)")+
  xlab("Latitude (degrees)")+
  theme_bw()+
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust = 0.5))
plt
```

<!--
#### Log10 y-scale 
```{r fig12, fig.height = 8, out.width = '100%', echo=FALSE}
plt = ggplot(dfClean, aes(label, values))+
  geom_boxplot()+
  scale_y_log10()+
  ylab("Carbon (g/m^2/year)")+
  xlab("Latitude (degrees)")+
  theme_bw()+
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust = 0.5))
plt
```

## Boxplots of ocean GPP w/ hemispheres as factor

#### Standard y-scale
```{r fig13, fig.height = 8, out.width = '100%', echo=FALSE}
plt = ggplot(dfClean, aes(as.factor(midLatAbs), values, fill=hemi))+
  geom_boxplot(position=position_dodge(width=0.6))+
  ylab("Carbon (g/m^2/year)")+
  xlab("Latitude (degrees)")+
  guides(fill=guide_legend(title="Hemisphere"))+
  theme_bw()+
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust = 0.5))
plt
```

#### Log10 y-scale 
```{r fig14, fig.height = 8, out.width = '100%', echo=FALSE}
plt = ggplot(dfClean, aes(as.factor(midLatAbs), values, fill=hemi))+
  geom_boxplot(position=position_dodge(width=0.6))+
  scale_y_log10()+
  ylab("Carbon (g/m^2/year)")+
  xlab("Latitude (degrees)")+
  guides(fill=guide_legend(title="Hemisphere"))+
  theme_bw()+
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust = 0.5))
plt
```
-->


##### Fig 6. Map of GPP distribution
```{r fig15, out.width = '100%', echo=FALSE}
myTheme = rasterTheme(region = chlPal(255))
levelplot(ggpp, zscaleLog=TRUE, margin=F, par.settings = myTheme, maxpixels=1000000)
```

```{r makeDistDf, include=FALSE}
samp = sampleRandom(ggpp, 100000, na.rm=T, cells=T, xy=T)
ggppDist = as.data.frame(cbind(samp, dist=distR[samp[,1]]))
colnames(ggppDist) = c("cell", "x", "y", "ggpp", "dist")

# fresh water df
dfdistggpplat = data.frame(y=NA, ggpp=NA, dist=NA, midLat=NA)


for(i in 1:length(latInfo$midLat)){
  print(i)
  theseRows = filter(ggppDist, y <= latInfo$fromLat[i] & y > latInfo$toLat[i]) %>%
    select(y, ggpp, dist)
  if(nrow(theseRows) == 0){next}
  theseRows$midLat = rep(latInfo$midLat[i], nrow(theseRows))
  dfdistggpplat = rbind(dfdistggpplat, theseRows)
}

dfdistggpplatClean = dfdistggpplat[2:nrow(dfdistggpplat),]

dfdistggpplatClean$midLat = factor(dfdistggpplatClean$midLat, levels = latInfo$midLat)

```


Much of this spread is because productivity is so much higher near land than it is in the ocean. The map shows this well, as does the black dot plot (Figures 7, 8). This might play an important role in patterns of diadromy if some groups are only using the nearshore, whereas others migrate farther from shore. 



##### Fig 7. GPP by distance from land
```{r fig16, fig.height = 8, out.width = '100%', echo=FALSE}
plt = ggplot(ggppDist, aes(dist, ggpp))+
  geom_point(alpha=0.01)+
  ylab("Carbon (g/m^2/year)")+
  xlab("Degrees from Land (distance)")+
  theme_bw()
plt
```

##### Fig 8. GPP by distance from land for 5 degree latitude bins
```{r fig17, fig.height = 35, out.width = '100%', echo=FALSE}
plt = ggplot(dfdistggpplatClean, aes(dist, ggpp))+
  geom_point(alpha=0.05)+
  scale_y_log10()+
  facet_grid(midLat~.)+
  ylab("Carbon (g/m^2/year)")+
  xlab("Degrees from Land (distance)")+
  theme_bw()
plt
```